---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bean ðŸ«›

<!-- badges: start -->
[![R-CMD-check](https://github.com/paanwaris/bean/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/paanwaris/bean/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

## Ecological Motivation

The `bean` package provides a tool to address a fundamental challenge in species distribution modeling (SDM) and ecological niche modeling (ENM): **sampling bias**. Occurrence records for species are rarely collected through a systematic, random process. Instead, they often cluster in easily accessible areas (like roads and cities) or in well-studied research sites. This spatial bias can translate into an **environmental bias**, where the model incorrectly learns that the species is associated with the environmental conditions of those heavily sampled areas, rather than its true ecological requirements.

`bean` tackles this problem by thinning occurrence data in **environmental space**. The goal is to create a more uniform distribution of points across the species' observed environmental niche, reducing the influence of densely clustered records. This allows for the construction of a more accurate **fundamental niche** volume, which can then be projected into geographic space to create a less biased prediction of habitat suitability.

The name `bean` reflects the core principle of the method: ensuring that each "pod" (a grid cell in environmental space) contains only a specified number of "beans" (occurrence points).

## Installation

To use the package, you first need to install it from GitHub. The following code will check for the required `devtools` package, install it if necessary, and then install `bean`.

```{r, eval=FALSE}
# 1. Install devtools if you don't have it yet
if (!require("devtools")) {
  install.packages("devtools")
}

# 2. Install bean from GitHub
devtools::install_github("paanwaris/bean")
```

## The `bean` Protocol: A Step-by-Step Guide

The recommended workflow is a four-step process designed to be transparent and reproducible.

### Step 1: Data Preparation and Visualization

Before any analysis, it is crucial to load and inspect the data. This includes both the species occurrence data and the associated environmental variables (e.g., from WorldClim). The most critical pre-processing step is to remove any records with missing environmental data, as these cannot be used for thinning.

```{r setup}
# Load required libraries
library(bean)
library(dplyr)
library(ggplot2)

# Load the raw occurrence data
# This path might need to be adjusted based on your project structure.
occ_data_raw <- read.csv("data/P_maniculatus_samples.csv")

# --- Critical: Clean the data ---
# Remove rows with NA or other non-finite values in the environmental variables.
# The functions in 'bean' have internal checks, but it's best practice to do this explicitly.
occ_data <- occ_data_raw %>%
  filter(is.finite(BIO1) & is.finite(BIO12))

head(occ_data)

# Visualize the initial distribution in environmental space
ggplot(occ_data, aes(x = BIO1, y = BIO12)) +
  geom_point(alpha = 0.5, color = "darkred") +
  labs(
    title = "Original Occurrence Points in Environmental Space",
    subtitle = paste(nrow(occ_data), "total points (after cleaning)"),
    x = "Mean Annual Temperature (BIO1)",
    y = "Annual Precipitation (BIO12)",
    caption = "Data for Peromyscus maniculatus"
  ) +
  theme_bw()
```

### Step 2: Objective Grid Resolution using Pairwise Distances

The most critical parameter in environmental gridding is the `grid_resolution`. Instead of guessing this value, we can derive it objectively from the data by analyzing the **distribution of pairwise environmental distances**.

The logic is to calculate the Euclidean distance between all pairs of points in the (scaled) environmental space. A small quantile of this distribution (e.g., the 10th percentile) represents a typical distance between "close" points. Using this value as the grid resolution is statistically justified because it adapts the cell size to the inherent scale of clustering within the dataset.

The `find_env_resolution()` function automates this process.

```{r find-resolution}
# Set a seed for reproducibility of the resampling in the correlogram
set.seed(81)  

# Let's use the 10th percentile of distances as our resolution
resolution_results <- find_env_resolution(
  data = occ_data,
  env_vars = c("BIO1", "BIO12"),
  quantile = 0.1
)

# The function returns a suggested resolution and the full distance distribution
resolution_results

# We can also plot the distribution to visualize the analysis
# The blue line shows the distance at the chosen quantile.
plot(resolution_results)

# Let's use this objective resolution in the next step
grid_res <- resolution_results$suggested_resolution
```

### Step 3: Parameter Exploration with `find_optimal_cap()`

This is the most important step for ensuring a defensible thinning strategy. Instead of guessing parameters, `find_optimal_cap()` allows you to explore the trade-offs and make a data-driven choice.

**Key Parameters:**
* `grid_resolution`: This defines the size of the cells in your environmental grid. The choice is ecologically significant. A small value creates a fine grid, which is sensitive to small environmental variations but may not thin large, dense clusters effectively. A large value creates a coarse grid, which is better for thinning broad-scale bias but may group distinct environmental conditions together.
* `target_percent`: This is your goal for data retention. A value of `0.5` means you want to keep approximately 50% of your data.

The function returns two key recommendations to guide your choice:
1.  `best_cap_closest`: The cap that results in a point count *numerically closest* to your target.
2.  `best_cap_above_target`: The cap that results in a point count that is *closest to, but not below*, your target. This is often the safer, more conservative choice if you want to avoid losing too much data.

```{r find-and-thin-part1}
# For reproducibility of the random sampling within the function
set.seed(81) 

# You can manually define the grid resolution based on ecological knowledge
# grid_res <- 0.1 # A resolution of 0.1 unit for the environmental axies

# Let's target retaining 50% of the data
optimal_params <- find_optimal_cap(
  data = occ_data,
  env_vars = c("BIO1", "BIO12"),
  grid_resolution = grid_res,
  target_percent = 0.50
)

# The function automatically saves results to the output directory.
# We can also inspect the returned list object.
# Print the recommendations
optimal_params

# Visualize the search process to understand the trade-offs
# The plot is also saved as a PNG in the output directory.
plot(optimal_params)
#The plot and the output list show that to get closest to our target of 50%.
```


### Step 4: Apply Thinning 

#### Method A: Stochastic Thinning with `thin_env_density`

Based on the exploration in Step 2, you can now make an informed decision and apply the final thinning. For this protocol, we will proceed with the `best_cap_above_target` to ensure we meet our minimum data requirement.

```{r find-and-thin-part2}
# --- Choose a cap and apply thinning ---
# This logic ensures that even if one recommendation is NA, the code will not fail.

# Default to a safe value
chosen_cap <- optimal_params$best_cap_above_target
# Proceeding with cap 
chosen_cap

# Set a seed again for the final, reproducible thinning
set.seed(81) 

thinned_data <- thin_env_density(
  data = occ_data,
  env_vars = c("BIO1", "BIO12"),
  grid_resolution = grid_res, 
  max_per_cell = chosen_cap
)

thinned_data
```

### Visualize the Thinning Process with Grids

To see exactly what the function is doing, we can draw the environmental grid over our plots.

```{r calculate-grid-lines}
# --- Calculate Grid Line Positions ---
# The grid lines correspond to the 'grid_resolution' parameter.

# For the x-axis (BIO1)
x_range <- range(occ_data$BIO1, na.rm = TRUE)
x_breaks <- seq(
  from = floor(x_range[1] / grid_res) * grid_res,
  to = ceiling(x_range[2] / grid_res) * grid_res,
  by = grid_res
)

# For the y-axis (BIO12)
y_range <- range(occ_data$BIO12, na.rm = TRUE)
y_breaks <- seq(
  from = floor(y_range[1] / grid_res) * grid_res,
  to = ceiling(y_range[2] / grid_res) * grid_res,
  by = grid_res
)

# --- Create a Labeled Data Frame for Comparison ---
# We create a unique ID to robustly identify which points were kept.
occ_data_labeled <- occ_data %>%
  mutate(unique_id = paste(x, y, sep = "_"))

thinned_data_labeled <- thinned_data$thinned_data %>%
  mutate(unique_id = paste(x, y, sep = "_"))
```

Now, let's create the plots with the grid overlay.

#### Original Data with Grid

This plot shows the initial clustering of points within the environmental grid cells.

```{r plot-original-grid, fig.width=8, fig.height=6}
ggplot(occ_data, aes(x = BIO1, y = BIO12)) +
  geom_vline(xintercept = x_breaks, color = "grey70", linetype = "dashed", linewidth = 0.5) +
  geom_hline(yintercept = y_breaks, color = "grey70", linetype = "dashed", linewidth = 0.5) +
  geom_point(color = "#D55E00", alpha = 0.6, size = 1.5) +
  labs(
    title = "Original Points with Environmental Grid",
    subtitle = paste(nrow(occ_data), "total points"),
    x = "Mean Annual Temperature (BIO1)",
    y = "Annual Precipitation (BIO12)"
  ) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank())
```

#### Thinned Data with Grid

This plot shows the result: a maximum of `r chosen_cap` point(s) per cell.

```{r plot-thinned-grid, fig.width=8, fig.height=6}
ggplot(thinned_data$thinned_data, aes(x = BIO1, y = BIO12)) +
  geom_vline(xintercept = x_breaks, color = "grey70", linetype = "dashed", linewidth = 0.5) +
  geom_hline(yintercept = y_breaks, color = "grey70", linetype = "dashed", linewidth = 0.5) +
  geom_point(color = "#0072B2", alpha = 0.6, size = 1.5) +
  labs(
    title = "Thinned Occurrence Points on Environmental Grid",
    subtitle = paste(nrow(thinned_data), "points remaining (max", chosen_cap, "per cell)"),
    x = "Mean Annual Temperature (BIO1)",
    y = "Annual Precipitation (BIO12)"
  ) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank())
```

#### Combined Comparison with Grid

```{r plot-combined-comparison-grid, fig.width=8, fig.height=6}
ggplot() +
  # 1. Plot the original data as a background layer
  geom_point(data = occ_data, aes(x = BIO1, y = BIO12), 
             color = "#D55E00", alpha = 0.6, size = 2) +
  
  # 2. Add the grid lines
  geom_vline(xintercept = x_breaks, color = "grey70", linetype = "dashed", linewidth = 0.5) +
  geom_hline(yintercept = y_breaks, color = "grey70", linetype = "dashed", linewidth = 0.5) +
  
  # 3. Plot the thinned data on top in a prominent color
  geom_point(data = thinned_data$thinned_data, aes(x = BIO1, y = BIO12), 
             color = "#0072B2", alpha = 0.7, size = 1) +
  
  # 4. Add informative labels
  labs(
    title = "Thinned Points Overlaid on Original Data",
    subtitle = paste(nrow(thinned_data$thinned_data), "points remaining (blue) from", nrow(occ_data), "original points (orange)"),
    x = "Mean Annual Temperature (BIO1)",
    y = "Annual Precipitation (BIO12)"
  ) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank())
```

#### Method B: Deterministic Thinning with `thin_env_center`

This method is simpler as it does not require choosing a cap. It returns one point for every occupied grid cell.

```{r thin-center}
thinned_data_center <- thin_env_center(
  data = occ_data,
  env_vars = c("BIO1", "BIO12"),
  grid_resolution = grid_res
)

thinned_data_center
```

#### Create the Thinned Center Data with Grid
```{r}
ggplot() +
  # 1. Plot the original data points as a faded background
  geom_point(data = occ_data, 
             aes(x = BIO1, y = BIO12), 
             color = "grey40", alpha = 0.5, size = 1.5) + 
  # 2. Add the grid lines
  geom_vline(xintercept = x_breaks, color = "grey70", linetype = "dashed", linewidth = 0.5) +
  geom_hline(yintercept = y_breaks, color = "grey70", linetype = "dashed", linewidth = 0.5) +
  
  # 3. Plot the new grid cell centers on top
  geom_point(data = thinned_data_center$thinned_points, 
             aes(x = BIO1, y = BIO12), 
             color = "#D55E00", size = 2, shape = 3, stroke = 1) +
  
  # 4. Add informative labels
  labs(
    title = "Deterministic Thinning to Grid Cell Centers",
    subtitle = paste(nrow(thinned_data_center$thinned_points), 
                     "unique cell centers (orange crosses) from", 
                     nrow(thinned_data_center$original_points), 
                     "original points (grey)"),
    x = "Mean Annual Temperature (BIO1)",
    y = "Annual Precipitation (BIO12)"
  ) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank())
```

### Step 5: Delineate and Visualize the Niche Ellipse

The final step is to take the cleaned, thinned occurrence points and formalize the environmental niche by fitting a bivariate ellipse. The `fit_ellipsoid()` function delineates this niche boundary.

### Thinned Density Ellipsoid 

```{r fit-ellipse-part1}
# Fit an ellipse that contains 95% of the thinned data
niche_ellipse <- fit_ellipsoid(
  data = thinned_data$thinned_data,
  env_vars = c("BIO1", "BIO12"),
  level = 0.95
)

# The returned object contains all the details
# We can use the custom print() method for a clean summary
niche_ellipse

# And we can use the custom plot() method for a powerful visualization
plot(niche_ellipse)
```

### Thinned Center Ellipsoid 

```{r fit-ellipse-part2}
# Fit an ellipse that contains 95% of the thinned data
center_niche_ellipse <- fit_ellipsoid(
  data = thinned_data_center$thinned_points,
  env_vars = c("BIO1", "BIO12"),
  level = 0.95
)

# The returned object contains all the details
# We can use the custom print() method for a clean summary
center_niche_ellipse

# And we can use the custom plot() method for a powerful visualization
plot(center_niche_ellipse)
```

### The End â¤ï¸
