---
output: github_document
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bean 🫛

<!-- badges: start -->

[![R-CMD-check](https://github.com/paanwaris/bean/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/paanwaris/bean/actions/workflows/R-CMD-check.yaml)
[![Lifecycle:
experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![License:
MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)

<!-- badges: end -->

## Ecological Motivation

The `bean` package provides a tool to address a fundamental challenge in
species distribution modeling (SDM, or ecological niche modeling, ENM):
**sampling bias**. Occurrence records for species are rarely collected
through a systematic, stratified process. Instead, they often cluster in
easily accessible areas (like roads and cities) or in well-studied
research sites. This spatial bias can translate into an **environmental
bias**, where the model incorrectly learns that the species is
associated with the environmental conditions of those heavily sampled
areas, rather than its true ecological requirements.

`bean` tackles this problem by thinning occurrence data in
**environmental space**. The goal is to create a more uniform
distribution of points across the species' observed environmental niche,
reducing the influence of densely clustered records. This allows for the
construction of a more accurate **fundamental niche** volume, which can
then be projected into geographic space to create a less biased
prediction of area with environmental suitability.

The name `bean` reflects the core principle of the method: ensuring that
each "pod" (a grid cell in environmental space) contains only a
specified number of "beans" (occurrence points).

## Installation

To use the package, you first need to install it from GitHub. The
following code will check for the required `devtools` package, install
it if necessary, and then install `bean`.

```{r, eval=FALSE}
# 1. Install devtools if you don't have it yet
if (!require("devtools")) {
  install.packages("devtools")
}

# 2. Install bean from GitHub
devtools::install_github("paanwaris/bean")
```

## Package Loading

To perform the necessary analyses, we need several R packages for data
manipulation, visualization, and modeling.

```{r Installation, warning = FALSE, message = FALSE}
# Load required libraries
library(bean)
library(terra)
library(rgl)
```

## The `bean` Protocol: A Step-by-Step Guide

### Step 1: Data Preparation and Visualization

The first step is to load your raw species occurrence data and
environmental raster layers. It's always a good practice to visualize
the spatial distribution of your points and inspect the environmental
layers.

```{r setup-1, fig.width=10, fig.height=8}
# Load the raw occurrence data from the package
occ_file <- system.file("extdata", "Rusa_unicolor.csv", package = "bean")
occ_data_raw <- read.csv(occ_file)

# Display the first few rows to understand its structure
head(occ_data_raw)
```

```{r setup-2}
# Load the environmental raster layers
thai_env_file <- system.file("extdata", "thai_env.tif", package = "bean")
env <- terra::rast(c(thai_env_file))

# Plot the environmental layers to check their extent and values
plot(env, mar = c(1, 1, 2, 4))

# Visualize the spatial distribution of the occurrence points
ggplot(occ_data_raw, aes(x = x, y = y)) +
  geom_raster(data = as.data.frame(env[[1]], xy = TRUE), aes(x = x, y = y), fill = "gray80") +
  geom_point(alpha = 0.5, color = "darkred") +
  coord_fixed() +
  labs(title = "Sambar Deer Occurrence Points") +
  theme_classic()
```

### Step 2: Prepare Data for Environmental Thinning

Before any analysis, the raw data must be cleaned and standardized. The
`prepare_bean()` function streamlines this process by:

1.  Removing records with missing coordinates.

2.  Extracting environmental data for each point from the raster layers.

3.  Removing records that fall outside the raster extent.

This ensures all subsequent functions work with a clean, complete, and
scaled dataset.

```{r prepare-data}
# Run the preparation function to clean and scale the data
origin_dat_prepared <- prepare_bean(
  data = occ_data_raw,
  env_rasters = env,
  longitude = "x",
  latitude = "y",
  transform = "none"
)

# View the structure and summary of the clean, scaled data
head(origin_dat_prepared)
summary(origin_dat_prepared)

# Visualize the spatial distribution of the cleaned occurrence points
ggplot(origin_dat_prepared, aes(x = x, y = y)) +
  geom_raster(data = as.data.frame(env[[1]], xy = TRUE), aes(x = x, y = y), fill = "gray80") +
  geom_point(alpha = 0.5, color = "darkred") +
  coord_fixed() +
  labs(title = "Sambar Deer Occurrence Points (Cleaned)") +
  theme_classic()
```

### Step 3: Objective Grid Resolution using Nearest Neighbors

The most critical parameter in environmental gridding is the
`grid_resolution`. Instead of guessing this value, we can derive it
objectively from the data by analyzing the density of points in
environmental space.

The `find_env_resolution()` function uses a geometric "elbow" method
based on nearest-neighbor distances. This identifies the exact distance
where dense artificial clustering transitions into natural data spacing,
eliminating the need for arbitrary quantiles.

```{r find-resolution, fig.width=12, fig.height=6}
# Find the objective resolution using the elbow method
resolution_results <- find_env_resolution(
  data = origin_dat_prepared,
  env_vars = c("bio_1", "bio_4", "bio_12", "bio_15")
)

plot(resolution_results)

# Let's use this objective resolution in the next step
grid_res <- resolution_results$suggested_resolution
grid_res
```

### Step 4: Apply Thinning

Now that we have an objective, data-driven `grid_resolution`, we can
apply the thinning. We offer two methods: stochastic and deterministic.

#### Method A: Stochastic Thinning with `thin_env_nd`

This method randomly samples exactly one point from each occupied grid
cell, enforcing strict presence-only logic within the environmental
hypercube.

```{r find-and-thin-part2}
# Apply the stochastic thinning (using a seed for reproducibility)
thinned_stochastic <- thin_env_nd(
  data = origin_dat_prepared,
  env_vars = c("bio_1","bio_4", "bio_12", "bio_15"),
  grid_resolution = grid_res
)

# Print the summary of the thinning results
thinned_stochastic
head(thinned_stochastic$thinned_data)
```

#### Method B: Deterministic Thinning with `thin_env_center`

This method provides a simpler, non-random alternative. It returns a
single new point at the exact center of every occupied grid cell,
regardless of how many original points fell within it.

```{r thin-center}
# Apply the deterministic thinning
thinned_deterministic <- thin_env_center(
  data = origin_dat_prepared,
  env_vars = c("bio_1","bio_4", "bio_12", "bio_15"),
  grid_resolution = c(0.5,0.5,0.5,0.5)
)

# Print the summary of the thinning results
thinned_deterministic
head(thinned_deterministic$thinned_points)
```

### Step 5: Visualize the Thinning Results

The `plot_bean()` function provides a powerful way to visualize the
effect of thinning by overlaying the thinned points on the original data
within the environmental grid.

```{r plot-thinning-results, fig.width=10, fig.height=8}
# Visualize the stochastic thinning results
plot_bean(
  original_data = origin_dat_prepared,
  thinned_object = thinned_stochastic,
  env_vars = c("bio_1","bio_4", "bio_12", "bio_15")
)

# Visualize the deterministic thinning results
plot_bean(
  original_data = origin_dat_prepared,
  thinned_object = thinned_deterministic,
  env_vars = c("bio_1","bio_4", "bio_12", "bio_15")
)
```

```{r}
# Visualize the spatial distribution of the occurrence and thinned points
# Combine the data and add a label for the partition
plot_data <- dplyr::bind_rows(
  occ_data_raw %>% dplyr::mutate(Data_Type = "Original (Raw)"),
  thinned_stochastic$thinned_data %>% dplyr::mutate(Data_Type = "Thinned (Stochastic)")
)

# Lock the order so "Original" is always on the left partition
plot_data$Data_Type <- factor(plot_data$Data_Type, levels = c("Original (Raw)", "Thinned (Stochastic)"))

# Plot with a partition (facet_wrap)
ggplot(plot_data, aes(x = x, y = y)) +
  geom_raster(data = as.data.frame(env[[1]], xy = TRUE), aes(x = x, y = y), fill = "gray80") +
  geom_point(alpha = 0.5, color = "darkred") +
  coord_fixed() +
  facet_wrap(~Data_Type) + # <--- This creates the partition line!
  labs(title = "Spatial Point Distribution",
       x = "Longitude", y = "Latitude") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "grey90"),
        strip.text = element_text(face = "bold"))
```

### Step 6: Delineate and Visualize the Niche Ellipse

After thinning, we can formalize the environmental niche by fitting a
bivariate ellipse around the points. The `fit_ellipsoid()` function
delineates this boundary.

### Original Ellipsoid

```{r fit-ellipse-part1, rgl=TRUE, fig.cap="3D Stochastic Ellipsoid"}
# Fit an ellipse that contains 95% of the original data
origin_ellipse <- fit_ellipsoid(data = origin_dat_prepared, 
                                env_vars = c("bio_1","bio_4", "bio_12", "bio_15"), 
                                method = "covmat", 
                                level = 0.95)
# The returned object contains all the details
origin_ellipse

# And we can use the custom plot() method for a powerful visualization
plot(origin_ellipse, dims = c(1, 2))
# For interactive 3D: 
plot(origin_ellipse, dims = c(1, 2, 3))
```

### Stochastic Thinned Ellipsoid

```{r fit-ellipse-part2, rgl=TRUE, fig.cap="3D Stochastic Ellipsoid"}
# Fit an ellipse that contains 95% of the thinned data
stochastic_ellipse <- fit_ellipsoid(data = thinned_stochastic$thinned_data, 
                                    env_vars = c("bio_1","bio_4", "bio_12", "bio_15"), 
                                    method = "covmat", 
                                    level = 0.95)
stochastic_ellipse

plot(stochastic_ellipse, dims = c(1, 2))
# For interactive 3D: 
plot(stochastic_ellipse, dims = c(1, 2, 3))
```

### Deterministic Thinned Ellipsoid

```{r fit-ellipse-part3, rgl=TRUE, fig.cap="3D Deterministic Ellipsoid"}
# Fit an ellipse that contains 95% of the thinned data
deterministic_ellipse <- fit_ellipsoid(data = thinned_deterministic$thinned_points,
                                       env_vars = c("bio_1","bio_4", "bio_12", "bio_15"), 
                                       method = "covmat", 
                                       level = 0.95)
deterministic_ellipse

plot(deterministic_ellipse, dims = c(1, 2))
# For interactive 3D: 
plot(deterministic_ellipse, dims = c(1, 2, 3))
```

### Step 7: Predict Niche Suitability

Finally, we project the learned fundamental niche back into geographic
space. The `predict` method calculates the continuous Mahalanobis
distance and suitability scores across the landscape.

```{r}
# Predict using the unthinned ellipsoid
origin_pred <- predict(object = origin_ellipse,
                       newdata = env,
                       include_suitability = TRUE,
                       suitability_truncated = TRUE,
                       include_mahalanobis = TRUE,
                       mahalanobis_truncated = TRUE)
plot(origin_pred)
```

```{r}
# Predict using the stochastic thinned ellipsoid
stochastic_pred <- predict(object = stochastic_ellipse,
                           newdata = env,
                           include_suitability = TRUE,
                           suitability_truncated = TRUE,
                           include_mahalanobis = TRUE,
                           mahalanobis_truncated = TRUE)
plot(stochastic_pred)
```

```{r}
# Predict using the deterministic thinned ellipsoid
deterministic_pred <- predict(object = deterministic_ellipse,
                              newdata = env,
                              include_suitability = TRUE,
                              suitability_truncated = TRUE,
                              include_mahalanobis = TRUE,
                              mahalanobis_truncated = TRUE)
plot(deterministic_pred)
```

### The End ❤️
