---
title: "Data preparation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data preparation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 7,
  dpi = 100,
  out.width = "95%"
)
```

Load the package

```{r}
library(bean)
library(terra)
library(rgl)
library(ggplot2)
```

# Step 1: Import Data

The first step is to load your raw species occurrence data and
environmental raster layers. It's always a good practice to visualize
the spatial distribution of your points and inspect the environmental
layers.

```{r setup-1, fig.width=10, fig.height=8}
# Load the raw occurrence data from the package
occ_file <- system.file("extdata", "Rusa_unicolor.csv", package = "bean")
occ_data_raw <- read.csv(occ_file)

# Display the first few rows to understand its structure
head(occ_data_raw)
```

```{r setup-2}
# Load the environmental raster layers
thai_env_file <- system.file("extdata", "thai_env.tif", package = "bean")
env <- terra::rast(c(thai_env_file))

# Plot the environmental layers to check their extent and values
plot(env, mar = c(1, 1, 2, 4))

# Visualize the spatial distribution of the occurrence points
ggplot(occ_data_raw, aes(x = x, y = y)) +
  geom_raster(data = as.data.frame(env[[1]], xy = TRUE), aes(x = x, y = y), fill = "gray80") +
  geom_point(alpha = 0.5, color = "darkred") +
  coord_fixed() +
  labs(title = "Sambar Deer Occurrence Points") +
  theme_classic()
```

# Step 2: Prepare Data for Environmental Thinning

Before any analysis, the raw data must be cleaned and standardized. The
`prepare_bean()` function streamlines this process by:

1.  Removing records with missing coordinates.

2.  Extracting environmental data for each point from the raster layers.

3.  Removing records that fall outside the raster extent.

This ensures all subsequent functions work with a clean, complete, and
scaled dataset.

```{r prepare-data}
# Run the preparation function to clean and scale the data
origin_dat_prepared <- prepare_bean(
  data = occ_data_raw,
  env_rasters = env,
  longitude = "x",
  latitude = "y",
  transform = "none"
)

# View the structure and summary of the clean, scaled data
head(origin_dat_prepared)
summary(origin_dat_prepared)

# Visualize the spatial distribution of the cleaned occurrence points
ggplot(origin_dat_prepared, aes(x = x, y = y)) +
  geom_raster(data = as.data.frame(env[[1]], xy = TRUE), aes(x = x, y = y), fill = "gray80") +
  geom_point(alpha = 0.5, color = "darkred") +
  coord_fixed() +
  labs(title = "Sambar Deer Occurrence Points (Cleaned)") +
  theme_classic()
```
