---
title: "1. Data preparation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1. Data preparation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "90%"
)
has_terra   <- requireNamespace("terra",   quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

`prepare_bean()` cleans raw occurrence records and extracts environmental
values for each point:

1. drops records with missing coordinates;
2. (optionally) standardises the raster layers (`"scale"`) or projects them
   onto principal components (`"pca"`);
3. extracts environmental values for each occurrence;
4. drops records that fall outside the raster extent.

```{r, eval = has_terra}
library(bean)
library(terra)

occ_file <- system.file("extdata", "Rusa_unicolor.csv", package = "bean")
env_file <- system.file("extdata", "thai_env.tif",     package = "bean")

occ_data_raw <- read.csv(occ_file)
env <- terra::rast(env_file)
head(occ_data_raw)
```

Inspect spatial distribution (only if `ggplot2` is available):

```{r, eval = has_terra && has_ggplot2}
library(ggplot2)
env_df <- as.data.frame(env[[1]], xy = TRUE)
ggplot(occ_data_raw, aes(x, y)) +
  geom_raster(data = env_df, aes(x, y), fill = "grey80") +
  geom_point(alpha = 0.5, colour = "darkred") +
  coord_fixed() +
  theme_classic()
```

Run the preparation step:

```{r, eval = has_terra}
prepared <- prepare_bean(
  data        = occ_data_raw,
  env_rasters = env,
  longitude   = "x",
  latitude    = "y",
  transform   = "scale"
)
head(prepared)
```

For users without `terra`, the package ships a prepared copy of the same
data:

```{r}
data(origin_dat_prepared, package = "bean")
head(origin_dat_prepared)
```
