---
title: "Environmental thinning"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Environmental thinning}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 7,
  dpi = 100,
  out.width = "95%"
)
```

Load the package

```{r, results='hide', message=FALSE, warning=FALSE}
library(bean)
library(terra)
library(rgl)
```

# Step 3: Objective Grid Resolution using Nearest Neighbors

The most critical parameter in environmental gridding is the
`grid_resolution`. Instead of guessing this value, we can derive it
objectively from the data by analyzing the density of points in
environmental space.

The `find_env_resolution()` function uses a geometric "elbow" method
based on nearest-neighbor distances. This identifies the exact distance
where dense artificial clustering transitions into natural data spacing,
eliminating the need for arbitrary quantiles.

```{r find-resolution, fig.width=12, fig.height=6}
data(origin_dat_prepared, package = "bean")
# Find the objective resolution using the elbow method
resolution_results <- find_env_resolution(
  data = origin_dat_prepared,
  env_vars = c("bio_1", "bio_4", "bio_12", "bio_15")
)

plot(resolution_results)
```

```{r find-resolution, fig.width=12, fig.height=6}
# Let's use this objective resolution in the next step
grid_res <- resolution_results$suggested_resolution
grid_res
```

# Step 4: Apply Thinning

Now that we have an objective, data-driven `grid_resolution`, we can
apply the thinning. We offer two methods: stochastic and deterministic.

## Method A: Stochastic Thinning with `thin_env_nd`

This method randomly samples exactly one point from each occupied grid
cell, enforcing strict presence-only logic within the environmental
hypercube.

```{r find-and-thin-part2}
# Apply the stochastic thinning (using a seed for reproducibility)
thinned_stochastic <- thin_env_nd(
  data = origin_dat_prepared,
  env_vars = c("bio_1","bio_4", "bio_12", "bio_15"),
  grid_resolution = grid_res
)

# Print the summary of the thinning results
thinned_stochastic
head(thinned_stochastic$thinned_data)
```

## Method B: Deterministic Thinning with `thin_env_center`

This method provides a simpler, non-random alternative. It returns a
single new point at the exact center of every occupied grid cell,
regardless of how many original points fell within it.

```{r thin-center}
# Apply the deterministic thinning
thinned_deterministic <- thin_env_center(
  data = origin_dat_prepared,
  env_vars = c("bio_1","bio_4", "bio_12", "bio_15"),
  grid_resolution = c(0.5,0.5,0.5,0.5)
)

# Print the summary of the thinning results
thinned_deterministic
head(thinned_deterministic$thinned_points)
```

# Step 5: Visualize the Thinning Results

The `plot_bean()` function provides a powerful way to visualize the
effect of thinning by overlaying the thinned points on the original data
within the environmental grid.

```{r plot-thinning-results, fig.width=10, fig.height=8}
data(occ_data_raw, package = "bean")
data(origin_dat_prepared, package = "bean")
data(thinned_stochastic, package = "bean")
data(thinned_deterministic, package = "bean")

# Visualize the stochastic thinning results
plot_bean(
  original_data = origin_dat_prepared,
  thinned_object = thinned_stochastic,
  env_vars = c("bio_1","bio_4", "bio_12", "bio_15")
)

# Visualize the deterministic thinning results
plot_bean(
  original_data = origin_dat_prepared,
  thinned_object = thinned_deterministic,
  env_vars = c("bio_1","bio_4", "bio_12", "bio_15")
)
```

```{r}
# Visualize the spatial distribution of the occurrence and thinned points

# Load the environmental raster layers
thai_env_file <- system.file("extdata", "thai_env.tif", package = "bean")
env <- terra::rast(c(thai_env_file))

# Combine the data and add a label for the partition
plot_data <- dplyr::bind_rows(
  occ_data_raw %>% dplyr::mutate(Data_Type = "Original (Raw)"),
  thinned_stochastic$thinned_data %>% dplyr::mutate(Data_Type = "Thinned (Stochastic)")
)

# Lock the order so "Original" is always on the left partition
plot_data$Data_Type <- factor(plot_data$Data_Type, levels = c("Original (Raw)", "Thinned (Stochastic)"))

# Plot with a partition (facet_wrap)
ggplot(plot_data, aes(x = x, y = y)) +
  geom_raster(data = as.data.frame(env[[1]], xy = TRUE), aes(x = x, y = y), fill = "gray80") +
  geom_point(alpha = 0.5, color = "darkred") +
  coord_fixed() +
  facet_wrap(~Data_Type) + # <--- This creates the partition line!
  labs(title = "Spatial Point Distribution",
       x = "Longitude", y = "Latitude") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "grey90"),
        strip.text = element_text(face = "bold"))
```
