---
title: "2. Environmental thinning"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. Environmental thinning}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "90%"
)
```

```{r setup}
library(bean)
data(origin_dat_prepared, package = "bean")
env_vars <- c("bio_1", "bio_4", "bio_12", "bio_15")
```

## Choosing an objective grid resolution

`find_env_resolution()` selects a **kernel-density bandwidth** for each
environmental variable. The bandwidth is the scale at which the empirical
density of observations becomes smooth, and is a statistically defensible
choice for the edge length of an environmental grid cell. The default
selector is the Sheather–Jones plug-in estimator (Sheather & Jones, 1991);
Silverman's rule (`"silverman"`) and Scott's rule (`"scott"`) are also
available.

```{r}
res <- find_env_resolution(
  data     = origin_dat_prepared,
  env_vars = env_vars,
  method   = "sheather-jones"
)
res
```

Visualise the per-variable kernel density and the chosen bandwidth:

```{r, fig.width = 7, fig.height = 6}
plot(res)
```

## Stochastic thinning

`thin_env_nd()` randomly retains exactly one occurrence per occupied grid
cell. A `seed` makes the selection reproducible without disturbing the global
random state.

```{r}
thinned_stochastic <- thin_env_nd(
  data            = origin_dat_prepared,
  env_vars        = env_vars,
  grid_resolution = res$suggested_resolution,
  seed            = 1
)
thinned_stochastic
```

## Deterministic thinning

`thin_env_center()` replaces each occupied cell with a single point at the
geometric centre of the cell. There is no randomness, so the result is
exactly reproducible.

```{r}
thinned_deterministic <- thin_env_center(
  data            = origin_dat_prepared,
  env_vars        = env_vars,
  grid_resolution = res$suggested_resolution
)
thinned_deterministic
```

## Comparing original and thinned data

`plot_bean()` overlays the thinned points on the original cloud in a
scatterplot matrix.

```{r, fig.width = 7, fig.height = 7}
plot_bean(
  original_data  = origin_dat_prepared,
  thinned_object = thinned_stochastic,
  env_vars       = env_vars
)
```
