---
title: "Environmental thinning"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Environmental thinning}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 7,
  dpi = 100,
  out.width = "95%"
)
```

Load the package

```{r}
library(bean)
library(terra)
library(rgl)
```

# Step 3: Objective Grid Resolution using Nearest Neighbors

The most critical parameter in environmental gridding is the
`grid_resolution`. Instead of guessing this value, we can derive it
objectively from the data by analyzing the density of points in
environmental space.

The `find_env_resolution()` function uses a geometric "elbow" method
based on nearest-neighbor distances. This identifies the exact distance
where dense artificial clustering transitions into natural data spacing,
eliminating the need for arbitrary quantiles.

```{r find-resolution, fig.width=12, fig.height=6}
data(origin_dat_prepared, package = "bean")
# Find the objective resolution using the elbow method
resolution_results <- find_env_resolution(
  data = origin_dat_prepared,
  env_vars = c("bio_1", "bio_4", "bio_12", "bio_15")
)

plot(resolution_results)

# Let's use this objective resolution in the next step
grid_res <- resolution_results$suggested_resolution
grid_res
```

# Step 4: Apply Thinning

Now that we have an objective, data-driven `grid_resolution`, we can
apply the thinning. We offer two methods: stochastic and deterministic.

## Method A: Stochastic Thinning with `thin_env_nd`

This method randomly samples exactly one point from each occupied grid
cell, enforcing strict presence-only logic within the environmental
hypercube.

```{r find-and-thin-part2}
# Apply the stochastic thinning (using a seed for reproducibility)
thinned_stochastic <- thin_env_nd(
  data = origin_dat_prepared,
  env_vars = c("bio_1","bio_4", "bio_12", "bio_15"),
  grid_resolution = grid_res
)

# Print the summary of the thinning results
thinned_stochastic
head(thinned_stochastic$thinned_data)
```

## Method B: Deterministic Thinning with `thin_env_center`

This method provides a simpler, non-random alternative. It returns a
single new point at the exact center of every occupied grid cell,
regardless of how many original points fell within it.

```{r thin-center}
# Apply the deterministic thinning
thinned_deterministic <- thin_env_center(
  data = origin_dat_prepared,
  env_vars = c("bio_1","bio_4", "bio_12", "bio_15"),
  grid_resolution = c(0.5,0.5,0.5,0.5)
)

# Print the summary of the thinning results
thinned_deterministic
head(thinned_deterministic$thinned_points)
```
