---
title: "Niche modeling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Niche modeling}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 7,
  dpi = 100,
  out.width = "95%"
)
```

Load the package

```{r}
library(bean)
library(terra)
library(rgl)
library(ggplot2)
```

# Step 5: Visualize the Thinning Results

The `plot_bean()` function provides a powerful way to visualize the
effect of thinning by overlaying the thinned points on the original data
within the environmental grid.

```{r plot-thinning-results, fig.width=10, fig.height=8}
data(occ_data_raw, package = "bean")
data(origin_dat_prepared, package = "bean")
data(thinned_stochastic, package = "bean")
data(thinned_deterministic, package = "bean")

# Visualize the stochastic thinning results
plot_bean(
  original_data = origin_dat_prepared,
  thinned_object = thinned_stochastic,
  env_vars = c("bio_1","bio_4", "bio_12", "bio_15")
)

# Visualize the deterministic thinning results
plot_bean(
  original_data = origin_dat_prepared,
  thinned_object = thinned_deterministic,
  env_vars = c("bio_1","bio_4", "bio_12", "bio_15")
)
```

```{r}
# Visualize the spatial distribution of the occurrence and thinned points

# Load the environmental raster layers
thai_env_file <- system.file("extdata", "thai_env.tif", package = "bean")
env <- terra::rast(c(thai_env_file))

# Combine the data and add a label for the partition
plot_data <- dplyr::bind_rows(
  occ_data_raw %>% dplyr::mutate(Data_Type = "Original (Raw)"),
  thinned_stochastic$thinned_data %>% dplyr::mutate(Data_Type = "Thinned (Stochastic)")
)

# Lock the order so "Original" is always on the left partition
plot_data$Data_Type <- factor(plot_data$Data_Type, levels = c("Original (Raw)", "Thinned (Stochastic)"))

# Plot with a partition (facet_wrap)
ggplot(plot_data, aes(x = x, y = y)) +
  geom_raster(data = as.data.frame(env[[1]], xy = TRUE), aes(x = x, y = y), fill = "gray80") +
  geom_point(alpha = 0.5, color = "darkred") +
  coord_fixed() +
  facet_wrap(~Data_Type) + # <--- This creates the partition line!
  labs(title = "Spatial Point Distribution",
       x = "Longitude", y = "Latitude") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "grey90"),
        strip.text = element_text(face = "bold"))
```

# Step 6: Delineate and Visualize the Niche Ellipse

After thinning, we can formalize the environmental niche by fitting a
bivariate ellipse around the points. The `fit_ellipsoid()` function
delineates this boundary.

## Original Ellipsoid

```{r fit-ellipse-part1, rgl=TRUE, fig.cap="3D Stochastic Ellipsoid"}
# Fit an ellipse that contains 95% of the original data
origin_ellipse <- fit_ellipsoid(data = origin_dat_prepared, 
                                env_vars = c("bio_1","bio_4", "bio_12", "bio_15"), 
                                method = "covmat", 
                                level = 0.95)
# The returned object contains all the details
origin_ellipse

# And we can use the custom plot() method for a powerful visualization
plot(origin_ellipse, dims = c(1, 2))
# For interactive 3D: 
plot(origin_ellipse, dims = c(1, 2, 3))
```

## Stochastic Thinned Ellipsoid

```{r fit-ellipse-part2, rgl=TRUE, fig.cap="3D Stochastic Ellipsoid"}
# Fit an ellipse that contains 95% of the thinned data
stochastic_ellipse <- fit_ellipsoid(data = thinned_stochastic$thinned_data, 
                                    env_vars = c("bio_1","bio_4", "bio_12", "bio_15"), 
                                    method = "covmat", 
                                    level = 0.95)
stochastic_ellipse

plot(stochastic_ellipse, dims = c(1, 2))
# For interactive 3D: 
plot(stochastic_ellipse, dims = c(1, 2, 3))
```

## Deterministic Thinned Ellipsoid

```{r fit-ellipse-part3, rgl=TRUE, fig.cap="3D Deterministic Ellipsoid"}
# Fit an ellipse that contains 95% of the thinned data
deterministic_ellipse <- fit_ellipsoid(data = thinned_deterministic$thinned_points,
                                       env_vars = c("bio_1","bio_4", "bio_12", "bio_15"), 
                                       method = "covmat", 
                                       level = 0.95)
deterministic_ellipse

plot(deterministic_ellipse, dims = c(1, 2))
# For interactive 3D: 
plot(deterministic_ellipse, dims = c(1, 2, 3))
```

# Step 7: Predict Niche Suitability

Finally, we project the learned fundamental niche back into geographic
space. The `predict` method calculates the continuous Mahalanobis
distance and suitability scores across the landscape.

```{r}
# Predict using the unthinned ellipsoid
origin_pred <- predict(object = origin_ellipse,
                       newdata = env,
                       include_suitability = TRUE,
                       suitability_truncated = TRUE,
                       include_mahalanobis = TRUE,
                       mahalanobis_truncated = TRUE)
plot(origin_pred)
```

```{r}
# Predict using the stochastic thinned ellipsoid
stochastic_pred <- predict(object = stochastic_ellipse,
                           newdata = env,
                           include_suitability = TRUE,
                           suitability_truncated = TRUE,
                           include_mahalanobis = TRUE,
                           mahalanobis_truncated = TRUE)
plot(stochastic_pred)
```

```{r}
# Predict using the deterministic thinned ellipsoid
deterministic_pred <- predict(object = deterministic_ellipse,
                              newdata = env,
                              include_suitability = TRUE,
                              suitability_truncated = TRUE,
                              include_mahalanobis = TRUE,
                              mahalanobis_truncated = TRUE)
plot(deterministic_pred)
```
