---
title: "3. Niche modeling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. Niche modeling}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "90%"
)
has_terra <- requireNamespace("terra", quietly = TRUE)
```

After thinning, the environmental niche is summarised by a multivariate
ellipsoid via `fit_ellipsoid()`. Two estimators are available:

- `"covmat"` — classical sample mean and covariance;
- `"mve"` — robust Minimum Volume Ellipsoid (Rousseeuw, 1985).

A 95% chi-square contour is used by default; change it with `level`.

```{r setup}
library(bean)
data(origin_dat_prepared, package = "bean")
data(thinned_stochastic,  package = "bean")
data(thinned_deterministic, package = "bean")
env_vars <- c("bio_1", "bio_4", "bio_12", "bio_15")
```

## Fit an ellipsoid

```{r}
origin_ellipse <- fit_ellipsoid(
  data     = origin_dat_prepared,
  env_vars = env_vars,
  method   = "covmat",
  level    = 0.95
)
origin_ellipse
```

## Visualise (2-D)

```{r, fig.width = 6, fig.height = 6}
plot(origin_ellipse, dims = c("bio_1", "bio_12"))
```

For an interactive 3-D view, install the optional package **rgl** and call
`plot(origin_ellipse, dims = c(1, 2, 3))`. If `rgl` is not installed, the
function falls back to a 2-D plot of the first two requested variables.

## Compare thinned ellipsoids

```{r, fig.width = 6, fig.height = 6}
stochastic_ellipse <- fit_ellipsoid(
  data     = thinned_stochastic$thinned_data,
  env_vars = env_vars
)
plot(stochastic_ellipse, dims = c("bio_1", "bio_12"))

deterministic_ellipse <- fit_ellipsoid(
  data     = thinned_deterministic$thinned_points,
  env_vars = env_vars
)
plot(deterministic_ellipse, dims = c("bio_1", "bio_12"))
```

## Predict suitability on a raster

`predict()` returns Mahalanobis distance and a Gaussian suitability score.
It works on a `data.frame` or directly on a `terra::SpatRaster`.

```{r, eval = has_terra, fig.width = 7, fig.height = 5}
library(terra)
env <- terra::rast(system.file("extdata", "thai_env.tif", package = "bean"))

pred <- predict(
  object                 = origin_ellipse,
  newdata                = env,
  include_suitability    = TRUE,
  suitability_truncated  = TRUE,
  include_mahalanobis    = FALSE
)
plot(pred)
```
