---
title: "Niche modeling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Niche modeling}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 7,
  dpi = 100,
  out.width = "95%"
)
```

Load the package

```{r, results='hide', message=FALSE, warning=FALSE}
library(bean)
library(terra)
library(rgl)
library(ggplot2)
```

# Step 6: Delineate and Visualize the Niche Ellipse

After thinning, we can formalize the environmental niche by fitting a
bivariate ellipse around the points. The `fit_ellipsoid()` function
delineates this boundary.

## Original Ellipsoid

```{r fit-ellipse-part1, rgl=TRUE, fig.cap="3D Stochastic Ellipsoid"}
# Fit an ellipse that contains 95% of the original data
origin_ellipse <- fit_ellipsoid(data = origin_dat_prepared, 
                                env_vars = c("bio_1","bio_4", "bio_12", "bio_15"), 
                                method = "covmat", 
                                level = 0.95)
# The returned object contains all the details
origin_ellipse

# And we can use the custom plot() method for a powerful visualization
plot(origin_ellipse, dims = c(1, 2))
# For interactive 3D: 
plot(origin_ellipse, dims = c(1, 2, 3))
```

## Stochastic Thinned Ellipsoid

```{r fit-ellipse-part2, rgl=TRUE, fig.cap="3D Stochastic Ellipsoid"}
# Fit an ellipse that contains 95% of the thinned data
stochastic_ellipse <- fit_ellipsoid(data = thinned_stochastic$thinned_data, 
                                    env_vars = c("bio_1","bio_4", "bio_12", "bio_15"), 
                                    method = "covmat", 
                                    level = 0.95)
stochastic_ellipse

plot(stochastic_ellipse, dims = c(1, 2))
# For interactive 3D: 
plot(stochastic_ellipse, dims = c(1, 2, 3))
```

## Deterministic Thinned Ellipsoid

```{r fit-ellipse-part3, rgl=TRUE, fig.cap="3D Deterministic Ellipsoid"}
# Fit an ellipse that contains 95% of the thinned data
deterministic_ellipse <- fit_ellipsoid(data = thinned_deterministic$thinned_points,
                                       env_vars = c("bio_1","bio_4", "bio_12", "bio_15"), 
                                       method = "covmat", 
                                       level = 0.95)
deterministic_ellipse

plot(deterministic_ellipse, dims = c(1, 2))
# For interactive 3D: 
plot(deterministic_ellipse, dims = c(1, 2, 3))
```

# Step 7: Predict Niche Suitability

Finally, we project the learned fundamental niche back into geographic
space. The `predict` method calculates the continuous Mahalanobis
distance and suitability scores across the landscape.

```{r}
# Predict using the unthinned ellipsoid
origin_pred <- predict(object = origin_ellipse,
                       newdata = env,
                       include_suitability = TRUE,
                       suitability_truncated = TRUE,
                       include_mahalanobis = TRUE,
                       mahalanobis_truncated = TRUE)
plot(origin_pred)
```

```{r}
# Predict using the stochastic thinned ellipsoid
stochastic_pred <- predict(object = stochastic_ellipse,
                           newdata = env,
                           include_suitability = TRUE,
                           suitability_truncated = TRUE,
                           include_mahalanobis = TRUE,
                           mahalanobis_truncated = TRUE)
plot(stochastic_pred)
```

```{r}
# Predict using the deterministic thinned ellipsoid
deterministic_pred <- predict(object = deterministic_ellipse,
                              newdata = env,
                              include_suitability = TRUE,
                              suitability_truncated = TRUE,
                              include_mahalanobis = TRUE,
                              mahalanobis_truncated = TRUE)
plot(deterministic_pred)
```
